The Concurrent Calculi Formalisation Benchmark

### 2024-02-16 short-paper/short-paper.tex
Add some motivating related work in note form.

In this spirit, we propose a new collection of benchmark problems specifically designed to address typical issues that arise when mechanising formal models of concurrent and distributed programming languages in general, and process calculi in particular.
Concurrent calculi are notably subtle: for instance it took nine years before the incorrect proof of type safety in~\cite{Honda1998} was fixed in~\cite{Yoshida2007}.
Similarly, many key results in peer-reviewed papers on session types have turned out to be false later on~\cite{Gay2020,10.1145/3290343}, illustrating the value of machine-checkable proofs.
While results about concurrent formalisms have already been mechanised (as we will discuss further below), our experience is that choosing appropriate mechanisation techniques and tools remains a significant challenge and that the trade-offs are not well understood.
This often leads researchers toward a trial-and-error approach, resulting in sub-optimal solutions, wasted mechanisation effort, and techniques and results that are hard to reuse.
\cite{Cruz-Filipe2021b} notes that the high-level parts of mechanised proofs closely resemble the informal versions, and that the main challenge lies in getting the foundations right.

@@

\cite{Brady2017} notes that using algebraic effects is unwieldy when there are many communication channels and suggests an experimental approach using dependent uniqueness types.

@@

\cite{Maksimovic2015} does not have name restriction, thus dodging the challenge of scope extrusion.

@@

\cite{DBLP:conf/forte/ZalakainD21} notes that the combination of linearity and infinite behaviour is difficult since it requires uniquely determining the resources consumed by a process.

>>>>

In this spirit, we propose a new collection of benchmark problems specifically designed to address typical issues that arise when mechanising formal models of concurrent and distributed programming languages in general, and process calculi in particular.
Concurrent calculi are notably subtle: for instance it took nine years before the incorrect proof of type safety in the seminal paper~\cite{Honda1998} was fixed in~\cite{Yoshida2007}.
Similarly, many key results in peer-reviewed papers on session types have turned out to be false later on~\cite{Gay2020,10.1145/3290343}, illustrating the value of machine-checkable proofs.
While results about concurrent formalisms have already been mechanised (as we will discuss further below), our experience is that choosing appropriate mechanisation techniques and tools remains a significant challenge and that the trade-offs are not well understood.
This often leads researchers toward a trial-and-error approach, resulting in sub-optimal solutions, wasted mechanisation effort, and techniques and results that are hard to reuse.
\cite{Cruz-Filipe2021b} notes that the high-level parts of mechanised proofs closely resemble the informal versions, and that the main challenge lies in getting the foundations right.

@@

\cite{Brady2017} notes that using algebraic effects is unwieldy when there are many communication channels and suggests an experimental approach using dependent uniqueness types.
\cite{Zalakain2019} notes that systems lacking linear types must simulate linearity, and that using a linearity predicate complicates congruences.
They had to limit themselves to an asymmetric structural precongruence to be able to prove subject reduction.

@@

\cite{Maksimovic2015} does not have name restriction, thus dodging the challenge of scope extrusion.
\cite{Castro-Perez2021,Castro2020} note that popular presentations of session type systems are impossible to directly encode using intrinsically \(\alpha\)-convertible terms.
\cite{Castro2020} suggests using several binder constructs to solve this, but notes that this requires proving many variants of the same theorems.

@@

\cite{DBLP:conf/forte/ZalakainD21} notes that the combination of linearity and infinite behaviour is difficult since it requires uniquely determining the resources consumed by a process.
\cite{Zalakain2019} notes that mixing replication and linearity seems to be very difficult without using explicit type contexts.
